---
title: "An explanation of the code used for the Auparoparo Oreophryne taxonomy manuscript"
author: "Marguerite Butler"
date: "2026-08-08"
output: html_document
---

### Acknowledgements

Many thanks to David Wachsmuth for his [beautiful blog post](https://upgo.lab.mcgill.ca/2019/12/13/making-beautiful-maps/) that explains spatial data _so well_, along with how to make beautiful inset maps using `sf`, `ggplot2`, and `cowplot`. I ended up using `grid`ʻs `viewport`, but `cowplot` is great too. 

### Note 

This was an interesting experiment for me trying `quarto` for both integrating the code with documentation and writing the manuscript versus `R` plus `LaTeX`. The integration of the former is great, and helps with organization, but Iʻm sure that Iʻm happy with the loss of fine control that `LaTeX` offers. Also some of the `.docx` glitches drive me crazy - if what you saw in `.html` was what you got in `.pdf` and `.docx`, that might make it more worthwhile for me. Maybe Iʻll use `quarto` for code documentation and go back to `LaTeX` for the manuscript and tables?

# The code in this repository produces the figures (map and phylogeny) and tables in our manuscript.

 
The map provides the complete range for all known _Oreophryne_ sensu lato species, covering Papua New Guinea, Indonesia, and the Philippines. 

The phylogeny shows where the clades _Auparoparo_ and _Oreophryne_ fall within the subfamily Asterophryinae, along with support values at the nodes. 

All scripts should be run starting from the `Code` directory. 

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('maps.R')
```

# MAP (Figure 3)

## Setup for the mapping

The mapping script is in `maps.R`. Load the needed packages. `dplyr` is for data manipulation, `ggplot2` and `grid` are for plotting, and the remaining are map packages. The `sf` package is for mapping sf shape data, `ggspatial` adds annotations, and `rnaturalearth` provides the map data for countries. 

`sf` data refers to "simple features" data format, which is a standard for storing spatial data (i.e., representations of maps) together with non-spatial data (metadata). Read more about that [here](https://cran.r-project.org/web/packages/flexpolyline/vignettes/sf-support.html). 

```{r, packages}
#| message: FALSE
#| warning: FALSE
```

### Load IUCN redlist data, our metadata, save as sf

The [IUCN redlist](https://www.iucnredlist.org/) contains known ranges for species. We downloaded all _Oreophryne_ sensu lato data as shape files. We had to download the entire subfamily (Asterophryinae), which is quite a large shapefile, covert to dataframe, filter by genus, and convert back to shapefile. The filtered dataset is what we save for this repository, saved as `oreophryne_sensu_lato.rds`. 

```{r, shapedata}
```

Read in the metadata (a spreadsheet), some light filtering, and save as an _sf dataset. 

```{r, metadata}
```

### Point aesthetics

Choosing the colors and shapes for the points. (I left in a color palette that I found from a [wonderful blog post](https://upgo.lab.mcgill.ca/2019/12/13/making-beautiful-maps/) because itʻs pretty.)

```{r, pointspec}
```
### Define the viewport for the inset

The x and y limits for the map in Latitude and Longitude are set here. We will be zooming in with an inset with the limits specified here. 

We use the `viewport()` function from package `grid` to define a picture-in-picture region where a second plot can be plotted into. x and y specify the location of the viewport on the basemap (between 0 and 1). 

```{r, maplimitsinset}
```

## Creating the basemap 

We get the basemap from the `rnaturalearth` package, by specifying the country names and then zeroing in on the region of interest by specifying the latitude and longitude limits of the map. 

We set the water color to a pleasing light blue, and choose a light grey for the species range data so that it will have a good contrast with the white area (lacking _Oreophryne_ species, as far as we know), and the points that represent our specimens on the map. 

The function `make_basemap()` maps the correct region of the earth, sets our color scheme, and overlays the species distribution data as grey polygons. 

Note that the function is __pipeable__ with the map data piped into the function. This is specified by placing a "." in front of the map data ".map". All of the map functions will follow this convention so that we can pass the map data and string the functions together. 

```{r, map}
```


### Overlaying points on the basemap 

The grouping variable to indicate point aesthetics is `genus` from the `metadata`, along with the longitude and latitude for each point. The specific values for shape, fill, and size are mapped by the `scales_..._manual` in `geom_point` (we chose these values earlier).

The `map_points()` function adds the points and a legend to the map. We specify the placement of the legend in the lower left corner of the map. 

```{r, pointmap}
```

### Annotate the map

We can easily add a scale bar and north arrow using these handy functions from `ggspatial`.  

The second function, `zoom_box()`, draws a rectangle around the region we will zoom in on for the inset.

```{r, annotatedmap}
```

We can then easily create the basemap starting from the `map_sf` shapefile and piping the functions together. We save this object to write to file later.  

```{r, savebasemap}
#| message: FALSE
#| warning: FALSE
```

### Inset map

Making the inset map is then just a matter of reusing the functions to plot the same map with smaller map limits. We also tweak some of the options to make it look pretty, such as dropping the legend, integer y-limits, thicker white zoom box line, and smaller expansion around the map area. 

```{r, insetmap}
#| message: FALSE
#| warning: FALSE
```

### Putting the map together

The basemap and the inset map are now two map objects, so all we have to do is print the basemap first and the inset over it, while specifying to plot it in the `viewport`. 

```{r, printmap}
#| message: FALSE
#| warning: FALSE
```

### Write map to file

The final map is written to the `Results` directory as `fig3.map-dot.pdf` or `.png`.

```{r, writemap}
#| message: FALSE
#| warning: FALSE
```

# PHYLOGENY (Figure 2)

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('oreophylogeny.R')
```

## Setup

The phylogeny plotting script is in `oreophylogeny.R`. Load the needed packages. Additional packages used include `treeio` is for tree input and ouptut, `tidytree` and `ggtree` is for tree manipulation and plotting. 

See Guangchuang Yuʻs excellent book explaining his packages _Data Integration, Manipulation and Visualization of Phylogenetic Trees_ by <https://yulab-smu.top/treedata-book/index.html>

```{r, packages}
#| message: FALSE
#| warning: FALSE
```

Load my personal helper functions 

```{r, functions}
#| message: FALSE
#| warning: FALSE
```

Read in the files from the Data directory (in subdirectories as indicated) -- the tree file, the metadata, genus type species `genus_references.csv`, and color codes. 

```{r, loaddata}
```

## Plot the full tree 

The full tree is quite large and plotted to the `Products/Validation_Figures` directory so that we can check the data to make sure everything read in correctly. 

We use create a `ggtree` object and add tip labels and adjust line thickneses

```{r, plotfulltree}
#| message: FALSE
#| warning: FALSE
```

## Remove duplicate represenatatives of each species 

To improve readability, we drop some duplicate tips so that there is only one per species.  

We grab the species names from the tree, and use the `distinct` function by species name to get a list of tips to drop (and those to keep). Then use the `drop.tip` function to remove duplicates and the outgroup species, and plot the trimmed tree for checking. 

```{r, trimtree}
#| message: FALSE
#| warning: FALSE
```

## Final, pretty tree

Now we are ready to plot the final, pretty tree with _Auparoparo_ and _Oreophryne_ highlighted. First, we identify the node of the Most Recent Common Ancestor (MRCA) for each of these clades, using the `get_MRCA` function on the tip names, and associate the labels with the nodes. 

We can then create a `ggtree` and attatch the `MRCA_labels` data to the ggtree object which is used to annotate the node label (letter "A" or "B") for the MRCA nodes (the tree can be thought of as a spreadsheet with one row per node, and the MRCA_label data is just added as an additional column). 

IQTREE provides the boostrap support values for each node, and we convert these to color coded dots to indicate bootstrap support using hte `geom_point2` function,   

We use `geom_text2` function to add the `MRCA_labels` "A" and "B" on the respective nodes, and we can add the gradient on these clades when we print the final tree using the very nifty function `gradient_tree` together with the `print_output` function. 

```{r, fulltree}
#| message: FALSE
#| warning: FALSE
```

# TABLES

Letʻs face it, tables are a pain. We used the `flextable` package to make tables with tailored formatting for this manuscript, output to both quartoʻs native .html format which was great for collaborating, and to .docx format required for journal submission. 

`flextable` is very cool, and `quarto` integration is great, but I found at times that I missed the fine control that `LaTeX` allows. 

The inputs are in `../Data/Raw_data/` and the outputs are saved to `../Data/Processed_data/` if they are to be accessed by other code or markdown files, and to `../Products/Manuscript/Tables/` if they are final products for the manuscript. 

## Setup 

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('tables.R')
```

### Load the packages

Make sure you have these packages installed.

```{r, tablepackages}
```
### Standardize the look of the tables 

These useful functions store the settings that can be applied for each table just by calling the function. 

```{r, flextablesetup}
#| message: FALSE
#| warning: FALSE
```

## Locality table (Table 1)

Read in the raw data, and specify the column headers, caption, and formatting. Save table in `.docx` format and an R data object in case we want to use it again later in code or markdown.

```{r, localities}
#| message: FALSE
#| warning: FALSE
```

## Character State Definitions (Table 2)

Table explaining the characters that were analyzed and how they are defined. The table is set up similarly, but this table has a footer. 

```{r, characterdef}
#| message: FALSE
#| warning: FALSE
```

## Character Matrix (Table 3)

The table reporting the character values for each species assessed. This table involves more filtering and formatting to fit on a page. This table also includes footnotes in particular cells/headers.

```{r, charactermatrix}
#| message: FALSE
#| warning: FALSE
```

## Reference Species (Table 4)

Table of reference species (type species or proxy) for each genus-level clade used in the phylogenetic definition of the clades. 

```{r, genustypes}
#| message: FALSE
#| warning: FALSE
```

## Taxonomic Transfers (Supplementary table 2)

The taxonomy table, listing all known species of _Oreophryne_ as currently recognized and to which genus it can be assigned.  Some filtering (separating words, italics, etc.) and light editing of the data (correcting the gender, etc.).  

```{r, taxonomy}
#| message: FALSE
#| warning: FALSE
```

## Type Localties 

This is a table made for our convenience to compile the data found in many papers.

```{r, typelocalities}
#| message: FALSE
#| warning: FALSE
```



